Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans

### 2020-06-03 paper_PLOS/paper_PLOS_draft.tex
Refactored figures and  tables, fixed Figure 6, broke Figure 4&7.

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}

@@

)

@@

One of the major constraints on the evolution of large body sizes in
animals is an increased risk of developing cancer. If all cells in all
organisms have a similar risk of malignant transformation and equivalent
cancer suppression mechanisms, organism with many cells should have a
higher prevalence of cancer than organisms with fewer cells. Consistent
with this expectation there is a strong positive correlation between
body size and cancer incidence within species, for example, human cancer
incidence increases with increasing adult height {[}1,2{]} and cancer
incidence is positively correlated with body size in dogs {[}3,4{]}.
There is no correlation, however, between body size and cancer risk
between species. This lack of correlation is often referred to as
`Peto's Paradox' {[}5--7{]}. While it is clear that a resolution to
Peto's Paradox must involve the evolution of enhanced cancer protection
alongside increases in body size and lifespan, the specific genetic,
molecular, and cellular mechanisms that underlie this resistance have
proven elusive. {[}8--12{]}.

Among the challenges for discovering how animals evolved enhanced cancer
protection mechanisms is identifying lineages in which large bodied
species are nested within species with small body sizes. Afrotherian
mammals are generally small-bodied, similarly to the prediced common
ancestor of Eutherian mammals. For example, maximum adult weights are
\textasciitilde{}70g in golden moles, \textasciitilde{}120g in tenrecs,
\textasciitilde{}170g in elephant shrews, \textasciitilde{}3kg in
hyraxes, and 60kg in aardvarks {[}13{]}. However, while these extant
species are relatively small, the fossil evidence demonstrates that
their ancestral lineages reached enormous sizes. For example, while
extant hyraxes are relatively small, the extinct Titanohyrax is
estimated to have weighted up to \textasciitilde{}1300kg {[}14{]}. The
largest members of Afrotheria, too, are dwarfed by the size of their
recent ancestors: extant cows manatees are large bodied
(\textasciitilde{}322-480kg) but are relatively small compared to the
extinct Stellar's sea cow which is estimated to have weight 8000-10000kg
{[}15{]}. Similarly African (4,800kg) and Asian elephants (3,200kg) are
the largest living elephant species, but are dwarfed by the truly
gigantic extinct Proboscideans such as Deinotherium
(\textasciitilde{}132,000kg), Mammut borsoni (110,000kg), and the Asian
straight-tusked elephant (\textasciitilde{}220,000kg), the largest known
land mammal {[}16{]}. These large-bodied Afrotherian lineages are nested
within small bodied species (Fig. 1) {[}17--20{]}, indicating that
gigantism independently evolved in hyraxes, sea cows, and elephants
(Paenungulates). Thus, Paenungulates are an excellent model system in
which to explore the mechanisms that underlie the evolution of large
body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto's paradox, among
the most parsimonious routes to enhanced cancer resistance is through an
increased copy number of tumor suppressors. Such an example has been
seen in the case of candidate genes such as \emph{TP53} and \emph{LIF}
{[}12,21,22{]} as well as in studies involving a limited set of
candidate genes {[}23,24{]}. As these studies focus on \emph{a priori}
gene sets, however, it remains unknown whether this is a general,
genome-wide trend in Afrotherian genomes; and whether such a general
trend is associated with the recent increases in body size -- and
therefore expected cancer risk -- in these species.

Here, we trace the evolution of body mass and gene copy number variation
in Afrotherians in order to investigate whether gene duplications are
enriched in large, long-lived species for genes involved in known tumor
suppression pathways. Our estimates of the evolution of body mass,
similarly to previous studies {[}17--20{]}, show that large body masses
evolved in a step-wise manner, with major increases in body mass in the
Pseudoungulata (17kg), Paenungulata (25kg), Tethytheria (296kg), and
Proboscidea (4,100kg) stem-lineages. Furthermore, we see that the
ancestral body size increases in Hydracoidia and Sirenia were
independent events. To study the evolution of gene copy number, we used
a genome-wide Reciprocal Best BLAT Hit (RBBH) method to identify gene
duplications in Afrotherian genomes, and used maximum likelihood
(treating copy number as a discrete trait) to infer the lineages in
which those duplications occurred. We found gene duplications in
lineages with increased body mass were enriched in functions related to
tumor suppression, including regulation of the cell cycle, DNA damage
repair, and regulation of apoptosis. These data suggest that duplication
of tumor suppressors played a role in the evolution of large, long-lived
in Afrotherians.

@@

We built a time-calibrated supertree of Eutherian mammals by combining
the time-calibrated molecular phylogeny of Bininda-Emonds \emph{et al.}
{[}25{]} with the time-calibrated total evidence Afrotherian phylogeny
from Puttick and Thomas {[}20{]}. While the Bininda-Emonds \emph{et al.}
{[}25{]} phylogeny includes 1,679 species, only 34 are Afrotherian, and
no fossil data are included. The inclusion of fossil data from extinct
species is essential to ensure that ancestral state reconstructions of
body mass are not biased by only including extant species. This can lead
to inaccurate reconstructions, for example, if lineages convergently
evolved large body masses from a small bodied ancestor. In contrast, the
total evidence Afrotherian phylogeny of Puttick and Thomas {[}20{]}
includes 77 extant species and fossil data from 39 extinct species.
Therefore we replaced the Afrotherian clade in the Bininda-Emonds
\emph{et al.} {[}25{]} phylogeny with the Afrotherian phylogeny of
Puttick and Thomas {[}20{]} using Mesquite. Next, we jointly estimated
rates of body mass evolution and reconstructed ancestral states using a
generalization of the Brownian motion model that relaxes assumptions of
neutrality and gradualism by considering increments to evolving
characters to be drawn from a heavy-tailed stable distribution (the
``Stable Model'') {[}26{]}. The stable model allows for occasional large
jumps in traits and has previously been shown to out-perform other
models of body mass evolution, including standard Brownian motion
models, Ornstein--Uhlenbeck models, early burst maximum likelihood
models, and heterogeneous multi-rate models {[}26{]}.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit
BLAT (RBHB) pipeline to identify putative homologs and estimate gene
copy numbers across species (\textbf{Figure 1A}). The Reciprocal Best
Hit (RBH) search strategy is conceptually straightforward: 1) Given a
gene of interest \(G_A\) in a query genome \(A\), one searches a target
genome \(B\) for all possible matches to \(G_A\); 2) For each of these
hits, one then performs the reciprocal search in the original query
genome to identify the highest-scoring hit; 3) A hit in genome \(B\) is
defined as a homolog of gene \(G_A\) if and only if the original gene
\(G_A\) is the top reciprocal search hit in genome \(A\). We selected
BLAT {[}27{]} as our algorithm of choice, as this algorithm is sensitive
to highly simliar (\textgreater{}90\% identity) sequences, thus
identifying the highest-confidence homologs while minimizing many-to-one
mapping problems when searching for multiple genes. RBH performs similar
to other more complex methods of orthology prediction, and is
particularly good at identifying incomplete genes that may be fragmented
in low quality/poor assembled regions of the genome {[}28,29{]}.

\textbf{Effective Copy Number By Coverage:} In lower-quality genomes,
many genes are fragmented across multiple scaffolds, which results in
BLAT calling multiple hits when in reality there is only one gene. To
compensate for this, we came up with a novel statistic, Estimated Copy
Number by Coverage (ECNC), which averages the number of times we see
each nucleotides of a query sequence in a target genome over the total
number of nucleotides of the query sequence found overall in each target
genome (Supplementary Figure 1). This allows us to correct for genes
that have been fragmented across incomplete genomes, while also taking
into account missing sequences from the human query in the target
genome. Mathematically, this can be written as:\\
\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\] where
\(n\) is a given nucleotide in the query, \(l\) is the total length of
the query, \(C_n\) is the number of instances that \(n\) is present
within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0
if \(C_n = 0\).

\textbf{RecSearch Pipeline:} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}30{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits.

After filtering out problematic protein queries (see ``Query gene
inclusion criteria''), we then used our pipeline (Figure 1A) to search
for all copies of our 18011 query genes in publicly available
Afrotherian genomes, including African savannah elephant
(\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African forest
elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\textbf{Query gene inclusion criteria:} To assemble our query list, we
first removed all unnamed genes from UP000005640. Next, we excluded
genes from downstream analyses for which assignment of homology was
uncertain, including uncharacterized ORFs (991), LOC (63), HLA genes
(402), replication dependent histones (72), odorant receptors (499),
ribosomal proteins (410), zinc finger transcription factors (1983),
viral and repetitive-element-associated proteins (82) and any protein
described as either ``Uncharacterized,'' ``Putative,'' or ``Fragment''
by UniProt in UP000005640 (30724), leaving us with a final set of 37582
query protein sequences, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria:} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes {[}12,22{]},
therefore partial gene duplications were included provided they passed
additional inclusion criteria. The ECNC cut off of 1.5 was selected
empirically, as this value minimized the number of false positives seen
in a test set of genes and genomes. The results of our initial search
are summarized in Figure 1B. Overall, we identified 13880 genes across
all species, or 77.1\% of our starting query genes.

\textbf{Genome Quality Assessment using CEGMA:} In order to determine
the effect of genome quality on our results, we used the gVolante
webserver and CEGMA to assess the quality and completeness of the genome
{[}31,32{]}. CEGMA was run using the default settings for mammals
(``Cut-off length for sequence statistics and composition'' = 1;``CEGMA
max intron length'' = 100000; ``CEGMA gene flanks'' = 10000, ``Selected
reference gene set'' = CVG). For each genome, we generated a correlation
matrix using the aforementioned genome quality scores, and either the
mean Copy Number or mean ECNC for all hits in the genome.

\hypertarget{evidence-for-functionality-of-identified-genes}{%
\subsection{Evidence for Functionality of Identified
Genes}\label{evidence-for-functionality-of-identified-genes}}

@@

\textbf{Transcriptome Assembly:} For the African Savana Elephant, Asian
Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated
\emph{de novo} transcriptomes using publically-available RNA-sequencing
data from NCBI SRA (Supplementary Table 1). We mapped reads to all
genomes available for each species, and assembled transcripts using
HISAT2 and StringTie, respectively {[}33--35{]}. RNA-sequencing data was
not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax,
Aardvark, or the Lesser Hedgehog Tenrec.

@@

We implemented a maximum likelihood method for determining the ancestral
copy numbers of genes in \emph{Atlantogenata} using IQ-Tree. For this
analysis, we used an unrooted species tree including only the
aforementioned \emph{Atlantogenata} species. We generated PHYLIP files
containing the copy number of each gene in the highest quality genome
for each species, encoding genes on a scale from 1-31+ copies as 1-9,
A-V; and encoding a gene's copy number as uncetain (``?'') when we did
not identify it in the genome. We used the included tree-searching and
model-testing functionality in IQ-Tree to determine the most likely
topology for the species tree, and to obtain the most likely model for
copy number changes in the genome. The most likely model for the
evolution of copy number was inferred to be a Jukes-Cantor type model
for morphological data with equal state frequencies, allowing for a
proportion of invariable sites, and using a discrete Gamma model with 4
rate categories. (``MK+FQ+I+G4''). We defined the ancestral state of a
node if it had greater than an 80\% posterior probability.

@@

To determine which pathways were associated with duplicated genes in
each species and lineage, we used WEBGESTALT to perform
overrepresentation analysis (ORA) of the duplicated gene lists relative
to our initial query gene list {[}36{]}. For the database of pathways
used in the analysis, we used Reactome {[}37{]}, Wikipathways, {[}38{]},
and KEGG {[}39{]}. For the ORA, we used FDR for determining
significance, and ran the analysis at FDR=0.1, FDR=0.2, FDR=0.3, and
FDR=0.5.

\hypertarget{lifespan-phylogenetic-least-square-regression-and-calculating-estimated-cancer-risk-thoughout-atlantogenata}{%
\subsection{Lifespan Phylogenetic Least-Square Regression and
Calculating Estimated Cancer Risk Thoughout
Atlantogenata}\label{lifespan-phylogenetic-least-square-regression-and-calculating-estimated-cancer-risk-thoughout-atlantogenata}}

In order to determine the cancer risk \(K\) at each node, we first
needed to calcultate ancestral lifespans at each node. To do so, we used
a Phylogenetic Generalized Least-Square Regression (PGLS) {[}40,41{]} to
calculate estimated ancestral lifespans across \emph{Atlantogenata}
using our estimates for body size at each node.

Next, we used a simplified multistage cancer risk model for body size
\(D\) and lifespan \(t\): \(K \approx Dt^6\) {[}7,42--44{]}. To
determine the change in \(K\) between nodes, we obtained the ratio
between the cancer risk \(K_{1}\) at any given node, and the cancer risk
\(K_{2}\) at its ancestral node, using the equation
\(\frac{K_{2}}{K_{1}} \approx \frac{D_{1}t_{1}^6}{D_{2}t_{2}^6}\).
Finally, to simplify comparisons, we calculated the fold change cancer
risk between a node and its ancestor as
\(\log_{2}(\frac{K_{2}}{K_{1}})\).

@@

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure1-1} \hfill{}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

@@

Among the major constraints on the evolution of large body sizes (and long life-spans) in animals is an increased risk of developing cancer. If all cells in all organisms have a similar risk of malignant transformation and equivalent cancer suppression mechanisms, organism with many cells should have a higher prevalence of cancer than organisms with fewer cells, particularly because large and small animals have similar cell sizes ({[}1{]}). Consistent with this expectation there is a strong positive correlation between body size and cancer incidence within species, for example, cancer incidence increases with increasing adult height in humans {[}2,3{]} and in dogs {[}4,5{]}. There is no correlation, however, between body size and cancer risk \emph{between} species; this lack of correlation is often referred to as `Peto's Paradox' {[}6--8{]}. The ultimate resolution to Peto's Paradox is obvious: large bodied and long-lived species evolved enhanced cancer protection mechanisms. However, identifiying the specific genetic, molecular, and cellular mechanisms that underlie the evolution of augmented cancer protection has been difficult {[}9--13{]}.

Among the challenges for discovering how animals evolved enhanced cancer protection mechanisms is identifying lineages in which large bodied species are nested within species with small body sizes. \emph{Afrotherian} mammals are generally small-bodied, but also include the largest extant land mammals. For example, maximum adult weights are \textasciitilde{}70g in golden moles, \textasciitilde{}120g in tenrecs, \textasciitilde{}170g in elephant shrews, \textasciitilde{}3kg in hyraxes, and 60kg in aardvarks {[}14{]}. While extant hyraxes are relatively small, the extinct Titanohyrax is estimated to have weighted up to \textasciitilde{}1300kg {[}15{]}. The largest members of \emph{Afrotheria}, too, are dwarfed by the size of their recent ancestors: extant cows manatees are large bodied (\textasciitilde{}322-480kg) but are relatively small compared to the extinct Stellar's sea cow which is estimated to have weight 8000-10000kg {[}16{]}. Similarly African (4,800kg) and Asian elephants (3,200kg) are the largest living elephant species, but are dwarfed by the truly gigantic extinct Proboscideans such as Deinotherium (\textasciitilde{}132,000kg), Mammut borsoni (110,000kg), and the Asian straight-tusked elephant (\textasciitilde{}220,000kg), the largest known land mammal {[}17{]}. Remarkably, these large-bodied \emph{Afrotherian} lineages are nested within small bodied species (\textbf{Figure \ref{fig:Figure-intro}}) {[}18--21{]}, indicating that gigantism independently evolved in hyraxes, sea cows, and elephants (\emph{Paenungulata}). Thus, Paenungulates are an excellent model system in which to explore the mechanisms that underlie the evolution of large body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto's paradox, among the most parsimonious routes to enhanced cancer resistance is through an increased copy number of tumor suppressors. Indeed, candidate genes studies have found that the elephant genome encodes duplicate such as \emph{TP53} and \emph{LIF} {[}13,22,23{]} as well as other genes with putative tumor suppressive functions {[}24,25{]}. As these studies focus on \emph{a priori} gene sets, however, it remains unknown whether this is a general, genome-wide trend in \emph{Afrotherian} genomes; and whether such a general trend is associated with the recent increases in body size -- and therefore expected cancer risk -- in these species.

Here, we trace the evolution of body mass and gene copy number variation in across \emph{Afrotherian} genomes in order to investigate whether duplication of tumors suppresor is common in large, long-lived Proboscideans. Our estimates of the evolution of body mass, similarly to previous studies {[}18--21{]}, show that large body masses evolved in a step-wise manner, with major increases in body mass in the \emph{Pseudoungulata} (17kg), \emph{Paenungulata} (25kg), \emph{Tethytheria} (296kg), and \emph{Proboscidea} (4,100kg) stem-lineages. To explore whether duplication of tumor suppressor genes occurred coincident with the evolution of large body sizes, we used a genome-wide Reciprocal Best BLAT Hit (RBBH) strategy to identify gene duplications, and used maximum likelihood to infer the lineages in which those duplications occurred. Unexpectedly, we found that duplication of tumor suppressor genes was common in all Afrotherians, both large and small. These data suggest that duplication of tumor suppressor genes is pervasive in Afrotherians and proceeded the evolution of species with very large body sizes.

@@

We built a time-calibrated supertree of Eutherian mammals by combining the time-calibrated molecular phylogeny of Bininda-Emonds \emph{et al.} {[}26{]} with the time-calibrated total evidence Afrotherian phylogeny from Puttick and Thomas {[}21{]}. While the Bininda-Emonds \emph{et al.} {[}26{]} phylogeny includes 1,679 species, only 34 are Afrotherian, and no fossil data are included. The inclusion of fossil data from extinct species is essential to ensure that ancestral state reconstructions of body mass are not biased by only including extant species. This can lead to inaccurate reconstructions, for example, if lineages convergently evolved large body masses from a small bodied ancestor. In contrast, the total evidence Afrotherian phylogeny of Puttick and Thomas {[}21{]} includes 77 extant species and fossil data from 39 extinct species. Therefore we replaced the Afrotherian clade in the Bininda-Emonds \emph{et al.} {[}26{]} phylogeny with the Afrotherian phylogeny of Puttick and Thomas {[}21{]} using Mesquite. Next, we jointly estimated rates of body mass evolution and reconstructed ancestral states using a generalization of the Brownian motion model that relaxes assumptions of neutrality and gradualism by considering increments to evolving characters to be drawn from a heavy-tailed stable distribution (the ``Stable Model'') implemented in StableTraits {[}27{]}. The stable model allows for occasional large jumps in traits and has previously been shown to out-perform other models of body mass evolution, including standard Brownian motion models, Ornstein--Uhlenbeck models, early burst maximum likelihood models, and heterogeneous multi-rate models {[}27{]}.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit BLAT (RBHB) pipeline to identify putative homologs and estimate gene copy number across species (\ref{fig:Figure-RBHBStrategy}). The Reciprocal Best Hit (RBH) search strategy is conceptually straightforward: 1) Given a gene of interest \(G_A\) in a query genome \(A\), one searches a target genome \(B\) for all possible matches to \(G_A\); 2) For each of these hits, one then performs the reciprocal search in the original query genome to identify the highest-scoring hit; 3) A hit in genome \(B\) is defined as a homolog of gene \(G_A\) if and only if the original gene \(G_A\) is the top reciprocal search hit in genome \(A\). We selected BLAT {[}28{]} as our algorithm of choice, as this algorithm is sensitive to highly simliar (\textgreater{}90\% identity) sequences, thus identifying the highest-confidence homologs while minimizing many-to-one mapping problems when searching for multiple genes. RBH performs similar to other more complex methods of orthology prediction, and is particularly good at identifying incomplete genes that may be fragmented in low quality/poor assembled regions of the genome {[}29,30{]}.

\textbf{Effective Copy Number By Coverage:} In lower-quality genomes, many genes are fragmented across multiple scaffolds, which results in BLAT calling multiple hits when in reality there is only one gene. To compensate for this, we came up with a novel statistic, Estimated Copy Number by Coverage (ECNC), which averages the number of times we see each nucleotides of a query sequence in a target genome over the total number of nucleotides of the query sequence found overall in each target genome (\textbf{Figure \ref{fig:Figure-ECNC}}). This allows us to correct for genes that have been fragmented across incomplete genomes, while also taking into account missing sequences from the human query in the target genome. Mathematically, this can be written as:\\
\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\]
where \(n\) is a given nucleotide in the query, \(l\) is the total length of the query, \(C_n\) is the number of instances that \(n\) is present within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0 if \(C_n = 0\).

\textbf{RecSearch Pipeline:} We created a custom Python pipeline for automating RBHB searches between a single reference genome and multiple target genomes using a list of query sequences from the reference genome. For the query sequences in our search, we used the hg38 Proteome provided by UniProt {[}31{]}, which is a comprehensive set of protein sequences curated from a combination of predicted and validated protein sequences generated by the UniProt Consortium. In order to refine our search, we omitted protein sequences originating from long, noncoding RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open reading frames (C-ORFs); and sequences with highly repetitive sequences such as zinc fingers, protocadherins, and transposon-containing genes, as these were prone to high levels of false positive hits.

After filtering out problematic protein queries (see below), we then used our pipeline to search for all copies of our 18011 query genes in publicly available \emph{Afrotherian} genomes (\ref{tab:Table-Genomes}), including African savannah elephant (\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African forest elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant (\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}: mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1, triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1, oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}: echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3), Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2, choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}: chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}: eleEdw1).

\textbf{Query gene inclusion criteria:} To assemble our query list, we first removed all unnamed genes from UP000005640. Next, we excluded genes from downstream analyses for which assignment of homology was uncertain, including uncharacterized ORFs (991 genes), LOC (63 genes), HLA genes (402 genes), replication dependent histones (72 genes), odorant receptors (499 genes), ribosomal proteins (410 genes), zinc finger transcription factors (1983 genes), viral and repetitive-element-associated proteins (82 genes) and any protein described as either ``Uncharacterized,'' ``Putative,'' or ``Fragment'' by UniProt in UP000005640 (30724 genes), leaving us with a final set of 37582 query protein isoforms, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria:} In order to condense transcript-level hits into single gene loci, and to resolve many-to-one genome mappings, we removed exons where transcripts from different genes overlapped, and merged overlapping transcripts of the same gene into a single gene locus call. The resulting gene-level copy number table was then combined with the maximum ECNC values observed for each gene in order to call gene duplications. We called a gene duplicated if its copy number was two or more, and if the maximum ECNC value of all the gene transcripts searched was 1.5 or greater; previous studies have shown that incomplete duplications can encode functional genes {[}13,23{]}, therefore partial gene duplications were included provided they passed additional inclusion criteria (see below). The ECNC cut off of 1.5 was selected empirically, as this value minimized the number of false positives seen in a test set of genes and genomes. The results of our initial search are summarized in Figure \ref{fig:Figure-GeneDup-Cladogram}. Overall, we identified 13880 genes across all species, or 77.1\% of our starting query genes.

\textbf{Genome Quality Assessment using CEGMA:} In order to determine the effect of genome quality on our results, we used the gVolante webserver and CEGMA to assess the quality and completeness of the genome {[}32,33{]}. CEGMA was run using the default settings for mammals (``Cut-off length for sequence statistics and composition'' = 1;``CEGMA max intron length'' = 100000; ``CEGMA gene flanks'' = 10000, ``Selected reference gene set'' = CVG). For each genome, we generated a correlation matrix using the aforementioned genome quality scores, and either the mean Copy Number or mean ECNC for all hits in the genome.

\hypertarget{evidence-for-functionality-of-gene-duplicates}{%
\subsection{Evidence for Functionality of Gene Duplicates}\label{evidence-for-functionality-of-gene-duplicates}}

@@

\textbf{Transcriptome Assembly:} For the African Savana Elephant, Asian Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated \emph{de novo} transcriptomes using publically-available RNA-sequencing data from NCBI SRA (\ref{tab:Table-SRAData}). We mapped reads to all genomes available for each species, and assembled transcripts using HISAT2 and StringTie, respectively {[}34--36{]}. RNA-sequencing data was not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax, Aardvark, or the Lesser Hedgehog Tenrec.

@@

We encoded the copy number of each gene for each species as a discrete trait ranging from 0 (one gene copy) to 31 (for 32+ gene copies) and used IQ-TREE to select the best-fitting model of character evolution {[}37--41{]}, which was inferred to be a Jukes-Cantor type model for morphological data (MK) with equal character state frequencies (FQ) and rate heterogeneity across sites approximated by including a class of invariable sites (I) plus a discrete Gamma model with four rate categories (G4). Next we inferred gene duplication and loss events with the empirical Bayesian ancestral state reconstruction (ASR) method implemented in IQ-TREE {[}37--41{]}, the best fitting model of character evolution (MK+FQ+GR+I) {[}42,43{]}, and the unrooted species tree for Atlantogenata. We considered ancestral state reconstructions to be reliable if they had Bayeisan Posterior Probability (BPP) ?0.80; less reliable reconstructions were excluded for pathway analyses.

@@

To determine if gene duplications were enriched in particular biological pathways, we used the WEB-based Gene SeT AnaLysis Toolkit (WEBGESTALT){[}44{]} to perform overrepresentation analysis (ORA); pathway databases included Reactome {[}45{]}, Wikipathways, {[}46{]}, and KEGG {[}47{]}. Gene duplicates in each lineage were used as the foreground gene set, and the initial query set was used as the background gene set. Statistical significance of enriched terms was assessed with a hypergeometric test, and controlled for multiple testing using the Benjamini-Hochberg false discovery rate (FDR); each analysis was run at FDR=0.1, FDR=0.2, FDR=0.3, and FDR=0.5. In order to determine an empirical false positive rate for term enrichment, we randomly sampled 1000-10,000 genes from our background set 1000 times, and ran the aforementioned analyses to see which terms were likely to randomly appear; there were no terms which appeared at FDR \(\leq\) 0.3.

\hypertarget{estimating-the-evolution-of-cancer-risk}{%
\subsection{Estimating the Evolution of Cancer Risk}\label{estimating-the-evolution-of-cancer-risk}}

The dramatic increase in body mass and lifespan in some Afrotherian lineages implies those lineages must have also evolved reduced cancer risk. To infer the magnitude of these reductions we estimated differences in intrinsic cancer risk across extant and ancestral Afrotherians. Following Peto {[}48{]} we estimate the intrinsic cancer risk as the product of risk associated with body mass and lifespan. In order to determine the intrinsic cancer risk (\(K\)) across species and at ancestral nodes (see below), we first needed to estimate ancestral lifespans at each node. We used Phylogenetic Generalized Least-Square Regression (PGLS) {[}49,50{]} implemented in the R package \texttt{ape} to calculate estimated ancestral lifespans across \emph{Atlantogenata} using our estimates for body size at each node. In order to estimate the intrinsic cancer risk of a species, we first inferred lifespans at ancestral nodes using PGLS and the model \(ln(lifespan) = \beta_{1}corBrowninan + \beta_{2}ln(size) + \epsilon\). Next, we calculated \(K_{1}\) at all nodes, and then estimated the fold change in cancer susceptibility between ancestral and descendant nodes (Figure \ref{fig:Figure-CancerSuscept}, Table \ref{tab:Table-CancerSuscept}).A Next, in oder to calculate \(K_{1}\) at all nodes, we used a simplified multistage cancer risk model for body size \(D\) and lifespan \(t\): \(K \approx Dt^6\) {[}8,48,51,52{]}. The fold change in cancer risk between a node and its ancestor was then defined as \(\log_{2}(\frac{K_{2}}{K_{1}})\).

@@

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-intro-1} \hfill{}

\caption{Atlantogenatans with sequenced genomes, body sizes, and known lifespans [@PuttickAndThomas2015; @HAGR]}\label{fig:Figure-intro}
\end{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-BodySize-1} \hfill{}
